---
title: "650 Project: Investigating the relationship between body mass index and average systolic blood pressure, considering BMI’s interaction with race"
output: html_document
---

#Investigating the relationship between body mass index and average systolic blood pressure, considering BMI’s interaction with race

## Preparation
### Load Libraries and Data
```{r load libraries and data}
library.list<- c('ggplot2','readxl','dplyr','tidyverse','patchwork','lme4','effects','sjPlot',"emmeans","gridExtra") #load packages
sapply(library.list, library, character.only=T, quietly=T)
library(NHANES)
dat<-NHANES
```

### Variable of Interest and Data Exploration
We first explore our outcome by printing its quantiles. Then we filtered the original data set and includes only variables of interest. Finally, we select variables with no NA in any of the variables we are interested in.
```{r variables}
quantile(dat$BMI, na.rm = TRUE)
dat$bmi_quan<-ifelse(dat$BMI<21.58, 1,
                            ifelse(21.58<=dat$BMI & dat$BMI<25.98,2,
                                   ifelse(25.98<=dat$BMI & dat$BMI<30.89, 3,
                                   ifelse(dat$BMI>=30.89,4,NA))))
#select variables
dat2 <- dat %>% select(ID,BPSysAve,BMI,bmi_quan,Age,Gender,Race1,Education,DirectChol,Diabetes,PhysActiveDays,SmokeNow,TotChol,AlcoholYear)

#target_pop: nonmissing outcome and POI
target_pop<-dat2[!is.na(dat2$BPSysAve)&!is.na(dat2$BMI),]#8487
#target_pop: add incomplete column
target_pop$Incomplete = rep(0,nrow(target_pop))
for(i in 1:nrow(target_pop)){
  if(sum(is.na(target_pop[i,]))!=0){target_pop$Incomplete[i]=1}
}
table(target_pop$Incomplete)
#Complete cases for all variables
comp_cases<-target_pop[complete.cases(target_pop),]#1276
```

Here is a table comparing distribution of data in the overall data, the data with only complete cases, and data in each quantile of BMI.
```{r}
#install.packages("qwraps2")
library("qwraps2")
orig_opt <- options()$qwraps2_markup
options(qwraps2_markup = "markdown")
summarytab<-
  list("Average Systolic Blood Pressure" =
       list("mean (sd)" = ~ qwraps2::mean_sd(BPSysAve,na_rm=TRUE)),
       "BMI" =
       list("mean (sd)" = ~ qwraps2::mean_sd(BMI,na_rm=TRUE)),
       "Age" =
       list("min"       = ~ min(Age),
            "max"       = ~ max(Age),
            "mean (sd)" = ~ qwraps2::mean_sd(Age,na_rm=TRUE)),
       "Gender" =
       list("Male" = ~ qwraps2::n_perc0(Gender == "male",na_rm=TRUE),
            "Female"  = ~ qwraps2::n_perc0(Gender == "female",na_rm=TRUE)),
       "Race" =
       list("White" = ~ qwraps2::n_perc0(Race1 == "White",na_rm=TRUE),
            "Black"  = ~ qwraps2::n_perc0(Race1 == "Black",na_rm=TRUE),
            "Mexican"= ~ qwraps2::n_perc0(Race1 == "Mexican",na_rm=TRUE),
            "Hispanic"= ~ qwraps2::n_perc0(Race1 == "Hispanic",na_rm=TRUE),
            "Other" = ~ qwraps2::n_perc0(Race1 == "Other",na_rm=TRUE)),
        "Eduaction" =
       list("8th Grade" = ~ qwraps2::n_perc0(Education == "8th Grade",na_rm=TRUE),
            "9 - 11th Grade"  = ~ qwraps2::n_perc0(Education == "9 - 11th Grade",na_rm=TRUE),
            "High School"= ~ qwraps2::n_perc0(Education == "High School",na_rm=TRUE),
            "Some College"= ~ qwraps2::n_perc0(Education == "Some College",na_rm=TRUE),
            "College Grad" = ~ qwraps2::n_perc0(Education == "College Grad",na_rm=TRUE),
            "Missing" = ~ qwraps2::n_perc0(is.na(Education))),
       "HDL Cholesterol" =
       list("mean (sd)" = ~ qwraps2::mean_sd(DirectChol,na_rm=TRUE)),
       "Diabetes" =
       list("Has diabetes" = ~ qwraps2::n_perc0(Diabetes == "Yes",na_rm=TRUE),
            "Doesn't have diabetes"  = ~ qwraps2::n_perc0(Diabetes == "No",na_rm=TRUE),
            "Missing" = ~ qwraps2::n_perc0(is.na(Diabetes))),
       "Physical Activity Days per Week" =
       list("mean (sd)" = ~ qwraps2::mean_sd(PhysActiveDays,na_rm=TRUE)),
        "Smoking" =
       list("Smoker" = ~ qwraps2::n_perc0(SmokeNow == "Yes",na_rm=TRUE),
            "Nonsmoker"  = ~ qwraps2::n_perc0(SmokeNow == "No",na_rm=TRUE),
            "Missing" = ~ qwraps2::n_perc0(is.na(SmokeNow))),
        "Alcohol Use Days per year"=
            list("mean (sd)" = ~ qwraps2::mean_sd(AlcoholYear,na_rm=TRUE))
       )
#only for those with Systolic Blood Pressure 
overall_tot<-summary_table(target_pop,summarytab)
overall_comp<-summary_table(comp_cases,summarytab)
bmi_quan<-summary_table(comp_cases,summarytab, by=c("bmi_quan"))
comp_tab<-cbind(overall_tot,overall_comp,bmi_quan)
print(comp_tab,
      title = "Summary Statistics For Complete Cases",
      cnames = c("Overall", "Overall-Complete Cases(N=1276)","BMI Quantile 1", "BMI Quantile 2", "BMI Quantile 3","BMI Quantile 4"))
```

Now, lets look at for adults over 20, Missing vs complete cases.
```{r data with complete values}
total_adult<-target_pop[target_pop$Age>=20,]
overall_adult<-summary_table(total_adult,summarytab)
miss_comp<-summary_table(total_adult,summarytab, by=c("Incomplete"))
tab_miss<-cbind(overall_adult,miss_comp,"P-Value" = "")

# bmi_quan_tar<-summary_table(target_pop,summarytab, by=c("bmi_quan"))
pval_2 <-c(t.test(BPSysAve ~ Incomplete, data = total_adult)$p.value, 
          t.test(BMI ~ Incomplete, data = total_adult)$p.value, NA, NA,
              t.test(Age ~ Incomplete, data = total_adult)$p.value, 
              chisq.test(total_adult$Gender,total_adult$Incomplete)$p.value,NA,
              chisq.test(total_adult$Race1,total_adult$Incomplete)$p.value,NA,NA,NA,NA,
              chisq.test(total_adult$Education,total_adult$Incomplete)$p.value,NA,NA,NA,NA,NA,
              t.test(DirectChol ~ Incomplete, data = total_adult)$p.value,
              chisq.test(total_adult$Diabetes,total_adult$Incomplete)$p.value,NA,NA,
              t.test(PhysActiveDays ~ Incomplete, data = total_adult)$p.value,
              chisq.test(total_adult$SmokeNow,total_adult$Incomplete)$p.value, NA,NA,
              t.test(AlcoholYear ~ Incomplete, data = total_adult)$p.value)
pval_2ad<-p.adjust(pval_2, method = "BH")
pval_2ad<-round(pval_2ad,digits=2)
pval_2ad<-ifelse(is.na(pval_2ad), " ",ifelse(pval_2ad<0.001,"<0.001",pval_2ad))

tab_miss[82:108]<-pval_2ad
print(tab_miss,
      rtitle = "Summary Statistics",
      cnames = c("Total Cases(N=6,919)","Complete Cases(N=1276)","Incomplete Cases(N=5643)","P Values"))

```

## Model Diagnostics
### Unadjusted model: SBP ~ BMI
See from the plot below, we find there is a linear relationship between SBP and BMI, so we start our model from here.
```{r scatterplot SBPvsBMI}
comp_cases%>%ggplot(aes(x=BMI,
                        y=BPSysAve))+
              geom_point()
```

Fit an unadjusted model as below.
```{r SBPvsBMI model}
##Just Outcome and POI
model_1<-lm(BPSysAve~BMI,data=comp_cases)
summary(model_1)
tab_model(model_1)
```

There is a significant positive association between SBP and BMI in the unadjusted model.

### Adjusted Model
#### Add demographic variables
We first added the demographic variables.
```{r adjusted model}
#added the other covariates in groups, in a sequential pattern
#Demographic
comp_cases <- within(comp_cases, Race1 <- relevel(Race1, ref = "White"))#make white the reference group
model_2<-lm(BPSysAve~BMI+Age+Gender+Race1+Education,data=comp_cases)
summary(model_2)
tab_model(model_2)
```

BMI is still significant.

#### Add health-related confounders
Therefore, we include health confounders as well.
```{r add health variables}
#health variables:DirectChol+Diabetes+PhysActiveDays+SmokeNow
model_3<-lm(BPSysAve~BMI+Age+Gender+Race1+Education+
              DirectChol+Diabetes+PhysActiveDays+SmokeNow+AlcoholYear,data=comp_cases)
summary(model_3)
tab_model(model_3)

suppressMessages(plot_model(model=model_3,type="diag",show.data = TRUE))[1] #only interested in VIF plot
```

Again, BMI is still significant after adjusting for possible confounders. Note that other significant confounders:Age,Gender, Race1,Education,DirectChol.

## Adjusted Model with Interaction
Interaction terms: literature reviewed and some explorations
```{r}
#interaction terms:Literature reviewed
# Age*gender
comp_cases %>%
     ggplot(aes(x=Age, 
               y=BPSysAve, 
               color=Gender))+
     geom_smooth(method="lm")
# BMI*race
comp_cases %>%
     ggplot(aes(x=BMI, 
               y=BPSysAve, 
               color=Race1))+
     geom_smooth(method="lm",se = FALSE)
# Diabetes*cholesterol
comp_cases %>%
     ggplot(aes(x=DirectChol,
               y=BPSysAve,
               color=Diabetes))+
     geom_smooth(method="lm",se = FALSE)
#Gender*Race
interaction.plot(x.factor = comp_cases$Gender, 
                 trace.factor = comp_cases$Race1,
                 response = comp_cases$BPSysAve)
#Race*education
interaction.plot(x.factor = comp_cases$Race1, 
                 trace.factor = comp_cases$Education,
                 response = comp_cases$BPSysAve)
```

We found other significant interactions with race in the literature, so let's also add them in here. This is the final model.
```{r}
model_final<-lm(BPSysAve~BMI+
                  Age+Gender+Race1+Education+
                  DirectChol+Diabetes+PhysActiveDays+AlcoholYear+SmokeNow+
                  Age*Gender+BMI*Race1+DirectChol*Diabetes+Gender*Race1+Race1*Education,
                data=comp_cases)
summary(model_final)
tab_model(model_final)
plot_model(model_final,type="diag",show.data = TRUE)[2:4]
```

```{r}
library(sjPlot)
library(sjmisc)
library(sjlabelled)
tab_model(model_final) # print out the summary table
```

```{r}
#check whether there exists significant differences for adding an BMI:Race1 interaction: p = 0.233, not significant.
Null.mod<-lm(BPSysAve~BMI+
                  Age+Gender+Race1+Education+
                  DirectChol+Diabetes+PhysActiveDays+AlcoholYear+SmokeNow+
                  Age*Gender+DirectChol*Diabetes+Gender*Race1+Race1*Education,
                data=comp_cases)
anova(model_final,Null.mod)
#obtain the p-value of catagorical variables as a whole 
anova(model_final)
```

model diagnosis for final model:

LINE assumption:
```{r}
library(car)
avPlots(model_final) #use partial regression plots to check linearity 
```
```{r}
library(stats)
dat1 = comp_cases
dat1$stu_res = rstudent(model_final) #obtain externally studentized residuals 
Residuals = dat1$stu_res #rename externally studentized residuals
lag.plot(Residuals, main = "Lag-1 plot")  #check independence

plot_model(model_final,type="diag",show.data = TRUE)[2] #qq plot, check normality
plot_model(model_final,type="diag",show.data = TRUE)[3] #histogram, also check normality
plot_model(model_final,type="diag",show.data = TRUE)[4] #residual plot, check equal variance
```
multicolinearity:
```{r}
model_3<-lm(BPSysAve~BMI+Age+Gender+Race1+Education+
              DirectChol+Diabetes+PhysActiveDays+SmokeNow+AlcoholYear,data=comp_cases) # model_3 without interactoin
plot_model(model_3,type="diag",show.data = TRUE)[1] #obtain vif plot, check multicolinearity
```
influencial diagnostics:
```{r}
model.h = hatvalues(model_final) # leverage point
Hmisc::describe(model.h) # descriptive information
max(model.h)
#m.h[which(model.h>0.5)]#high leverage-0
model.h[which(model.h>0.2)]   #moderate leverage-14

dat_high_lev = dat1[c(138,139,256,271,287,414,690,940,958,1094,1095,1136,1167,1221),]
dat_high_lev # output moderate leverage points
```

```{r}
Hmisc::describe(dat1$stu_res) 
outlier_1 = dat1[abs(Residuals)>4,]  
as.data.frame(outlier_1) # obtain outliers with cut-off value = 4: 7 observations with ID 51933, 62727, 68301

car::influencePlot(model_final, main = "Regression influence plot") # obtain influential plot and influential points

olsrr::ols_plot_dffits(model_final) # DFFITS plot
olsrr::ols_plot_cooksd_chart(model_final) # Cook's distance plot
```
