---
title: "650 Project"
author: "Jess"
date: "11/28/2021"
output: html_document
---

```{r}
#load packages
#install.packages('ggplot2','dplyr','tidyverse','readxl','patchwork','lme4','effects',"devtools","emmeans","gridExtra") #install packages

library.list<- c('ggplot2','readxl','dplyr','tidyverse','patchwork','lme4','effects','sjPlot',"emmeans","gridExtra") #load packages
sapply(library.list, library, character.only=T, quietly=T)
library(NHANES)
dat<-NHANES
```

```{r}
quantile(dat$BMI, na.rm = TRUE)
dat$bmi_quan<-ifelse(dat$BMI<21.58, 1,
                            ifelse(21.58<=dat$BMI & dat$BMI<25.98,2,
                                   ifelse(25.98<=dat$BMI & dat$BMI<30.89, 3,
                                   ifelse(dat$BMI>=30.89,4,NA))))
#select variables
dat2 <- dat %>% select(ID,BPSysAve,BMI,bmi_quan,Age,Gender,Race1,Education,DirectChol,Diabetes,PhysActive,SmokeNow,AlcoholYear)

#target_pop: nonmissing outcome and POI
target_pop<-dat2[!is.na(dat2$BPSysAve)&!is.na(dat2$BMI),]#8487
#target_pop: add incomplete column
target_pop$Incomplete = rep(0,nrow(target_pop))
for(i in 1:nrow(target_pop)){
  if(sum(is.na(target_pop[i,]))!=0){target_pop$Incomplete[i]=1}
}
table(target_pop$Incomplete)
#Complete cases for all variables
comp_cases<-target_pop[complete.cases(target_pop),]#2636
```

## First lets look at a Table for the dataset we will be using (complete cases: comp_cases)
```{r}
#install.packages("qwraps2")
library("qwraps2")
orig_opt <- options()$qwraps2_markup
options(qwraps2_markup = "markdown")
summarytab<-
  list("Average Systolic Blood Pressure" =
       list("mean (sd)" = ~ qwraps2::mean_sd(BPSysAve,na_rm=TRUE)),
       "BMI" =
       list("mean (sd)" = ~ qwraps2::mean_sd(BMI,na_rm=TRUE)),
       "Age" =
       list("min"       = ~ min(Age),
            "max"       = ~ max(Age),
            "mean (sd)" = ~ qwraps2::mean_sd(Age,na_rm=TRUE)),
       "Gender" =
       list("Male" = ~ qwraps2::n_perc0(Gender == "male",na_rm=TRUE),
            "Female"  = ~ qwraps2::n_perc0(Gender == "female",na_rm=TRUE)),
       "Race" =
       list("White" = ~ qwraps2::n_perc0(Race1 == "White",na_rm=TRUE),
            "Black"  = ~ qwraps2::n_perc0(Race1 == "Black",na_rm=TRUE),
            "Mexican"= ~ qwraps2::n_perc0(Race1 == "Mexican",na_rm=TRUE),
            "Hispanic"= ~ qwraps2::n_perc0(Race1 == "Hispanic",na_rm=TRUE),
            "Other" = ~ qwraps2::n_perc0(Race1 == "Other",na_rm=TRUE),
            "Missing" = ~ qwraps2::n_perc0(is.na(Race1))),
        "Eduaction" =
       list("8th Grade" = ~ qwraps2::n_perc0(Education == "8th Grade",na_rm=TRUE),
            "9 - 11th Grade"  = ~ qwraps2::n_perc0(Education == "9 - 11th Grade",na_rm=TRUE),
            "High School"= ~ qwraps2::n_perc0(Education == "High School",na_rm=TRUE),
            "Some College"= ~ qwraps2::n_perc0(Education == "Some College",na_rm=TRUE),
            "College Grad" = ~ qwraps2::n_perc0(Education == "College Grad",na_rm=TRUE),
            "Missing" = ~ qwraps2::n_perc0(is.na(Education))),
       "HDL Cholesterol" =
       list("mean (sd)" = ~ qwraps2::mean_sd(DirectChol,na_rm=TRUE)),
       "Diabetes" =
       list("Has diabetes" = ~ qwraps2::n_perc0(Diabetes == "Yes",na_rm=TRUE),
            "Doesn't have diabetes"  = ~ qwraps2::n_perc0(Diabetes == "No",na_rm=TRUE),
            "Missing" = ~ qwraps2::n_perc0(is.na(Diabetes))),
       "Physical Activity" =
       list("Engages in moderate/intense exercise" = ~ qwraps2::n_perc0(PhysActive == "Yes",na_rm=TRUE),
            "Doesn't engages in moderate/intense exercise"  = ~ qwraps2::n_perc0(PhysActive == "No",na_rm=TRUE),
            "Missing" = ~ qwraps2::n_perc0(is.na(PhysActive))),
        "Smoking" =
       list("Smoker" = ~ qwraps2::n_perc0(SmokeNow == "Yes",na_rm=TRUE),
            "Nonsmoker"  = ~ qwraps2::n_perc0(SmokeNow == "No",na_rm=TRUE),
            "Missing" = ~ qwraps2::n_perc0(is.na(SmokeNow))),
        "Alcohol Use Days per year"=
            list("mean (sd)" = ~ qwraps2::mean_sd(AlcoholYear,na_rm=TRUE))
       )
#only for those with Systolic Blood Pressure 
overall_tot<-summary_table(target_pop,summarytab)
overall_comp<-summary_table(comp_cases,summarytab)
bmi_quan<-summary_table(comp_cases,summarytab, by=c("bmi_quan"))
comp_tab<-cbind(overall_tot,overall_comp,bmi_quan)
print(comp_tab,
      title = "Summary Statistics For Complete Cases",
      cnames = c("Overall", "Overall-Complete Cases(N=2636)","BMI Quartile 1", "BMI Quartile 2", "BMI Quartile 3","BMI Quartile 4"))
```
## Lets look at for adults over 20, Missing vs complete cases
```{r}
total_adult<-target_pop[target_pop$Age>=20,]
overall_adult<-summary_table(total_adult,summarytab)
miss_comp<-summary_table(total_adult,summarytab, by=c("Incomplete"))
tab_miss<-cbind(overall_adult,miss_comp,"P-Value" = "")

# bmi_quan_tar<-summary_table(target_pop,summarytab, by=c("bmi_quan"))
pval_2 <-c(t.test(BPSysAve ~ Incomplete, data = total_adult)$p.value, 
          t.test(BMI ~ Incomplete, data = total_adult)$p.value, NA, NA,
              t.test(Age ~ Incomplete, data = total_adult)$p.value, 
              chisq.test(total_adult$Gender,total_adult$Incomplete)$p.value,NA,
              chisq.test(total_adult$Race1,total_adult$Incomplete)$p.value,NA,NA,NA,NA,
              chisq.test(total_adult$Education,total_adult$Incomplete)$p.value,NA,NA,NA,NA,NA,
              t.test(DirectChol ~ Incomplete, data = total_adult)$p.value,
              chisq.test(total_adult$Diabetes,total_adult$Incomplete)$p.value,NA,NA,
              chisq.test(total_adult$PhysActive,total_adult$Incomplete)$p.value,
              chisq.test(total_adult$SmokeNow,total_adult$Incomplete)$p.value, NA,NA,
              t.test(AlcoholYear ~ Incomplete, data = total_adult)$p.value)
pval_2ad<-p.adjust(pval_2, method = "BH")
pval_2ad<-round(pval_2ad,digits=2)
pval_2ad<-ifelse(is.na(pval_2ad), " ",ifelse(pval_2ad<0.001,"<0.001",pval_2ad))

tab_miss[82:108]<-pval_2ad
print(tab_miss,
      rtitle = "Summary Statistics",
      cnames = c("Total Cases(N=8487)","Complete Cases(N=2636)","Incomplete Cases(N=5851)","P Values"))

```

## Lets look at SBP and BMI
```{r}
comp_cases%>%ggplot(aes(x=BMI,
                        y=BPSysAve))+
              geom_point()
```

## I then fit an unadjusted model
```{r}
##Just Outcome and POI
model_1<-lm(BPSysAve~BMI,data=comp_cases)
summary(model_1)
```
BMI is significant: there is an association between SBP and BMI

##I added the demographic variables next
```{r}
#I added the other covariates in groups, in a sequential pattern
#Demographic
comp_cases <- within(comp_cases, Race1 <- relevel(Race1, ref = "White"))#make white the reference group
model_2<-lm(BPSysAve~BMI+Gender+Race1+Education+Age,data=comp_cases)
summary(model_2)
```
#BMI is still significant.

##Lets include health confounders as well
```{r}
#health variables:DirectChol+Diabetes+PhysActive+SmokeNow
model_3<-lm(BPSysAve~BMI+Age+Gender+Race1+Education+
              DirectChol+Diabetes+PhysActive+SmokeNow+AlcoholYear,data=comp_cases)
summary(model_3)

plot_model(model_3,type="diag",show.data = TRUE) #only interested in VIF plot
```
Again, BMI is still significant. Other significant confounders:Age,Gender, Race1,Education,DirectChol

##Interaction terms: literature reviewed and some explorations
```{r}
#interaction terms:Literature reviewed
# Age*gender
comp_cases %>%
     ggplot(aes(x=Age, 
               y=BPSysAve, 
               color=Gender))+
     geom_smooth(method="lm")
# BMI*race
comp_cases %>%
     ggplot(aes(x=BMI, 
               y=BPSysAve, 
               color=Race1))+
     geom_smooth(method="lm",se = FALSE)
# Diabetes*cholesterol
comp_cases %>%
     ggplot(aes(x=DirectChol,
               y=BPSysAve,
               color=Diabetes))+
     geom_smooth(method="lm",se = FALSE)
#Gender*Race
interaction.plot(x.factor = comp_cases$Gender, 
                 trace.factor = comp_cases$Race1,
                 response = comp_cases$BPSysAve)
#Race*education
interaction.plot(x.factor = comp_cases$Race1, 
                 trace.factor = comp_cases$Education,
                 response = comp_cases$BPSysAve)
```

#We found other significant interactions with race in the literature, so lets also add them in here. This is the final model. 
```{r}
model_final<-lm(BPSysAve~BMI+
                  Age+Gender+Race1+Education+
                  DirectChol+Diabetes+PhysActive+AlcoholYear+SmokeNow+
                  Age*Gender+BMI*Race1+DirectChol*Diabetes+Gender*Race1+Race1*Education,
                data=comp_cases)
summary(model_final)

plot_model(model_final,type="diag",show.data = TRUE)
```
```{r}
#Is the interaction term significant
Null.mod<-lm(BPSysAve~BMI+
                  Age+Gender+Race1+Education+
                  DirectChol+Diabetes+PhysActive+AlcoholYear+SmokeNow+
                  Age*Gender+DirectChol*Diabetes+Gender*Race1+Race1*Education,
                data=comp_cases)
anova(model_final,Null.mod)
#are the categorical terms as a whole significant?
anova(model_final)
```

#My attempt at some diagnostics
```{r}
#install.packages("Hmisc")
m.h = hatvalues(model_final)
Hmisc::describe(m.h)

max(m.h)
#m.h[which(m.h>0.5)]#high leverage-0
m.h[which(m.h>0.2)]#moderate leverage-14
```

```{r}
#install.packages("olsrr")
options(width = 70)
options(warn=-1)

m.dffits=dffits(model_4)
m.dfbeta=dfbeta(model_4)
m.D=cooks.distance(model_4)
m.covratio=covratio(model_4)

n = nrow(comp_cases); p = model_4$rank
plot(model_4, which=4); abline(h=4/n,lty=2) #cooks difference: impact on the parameter estimates
plot(abs(m.covratio-1)); abline(1+3*p/n,0); abline(1-3*p/n,0) #influence on the precision of estimation of beta_hat
olsrr::ols_plot_dfbetas(model_4)
olsrr::ols_plot_dffits(model_4)
```

#partial regression plots
```{r}
car::avPlots(model_final)
car::residualPlots(model_final,type="response")
```

```{r}
p_unadj<-ifelse(summary(model_1)$coefficients[,"Pr(>|t|)"]<0.01,"<0.01",round(summary(model_1)$coefficients[,"Pr(>|t|)"],digits=2))
unadj_est<-summary(model_1)$coefficients[,"Estimate"]
unadj_se<-summary(model_1)$coefficients[,"Std. Error"]
unadj_nm<-rownames(summary(model_1)$coefficients)
unadjusted<-data.frame("Variable"=unadj_nm,"Estimates"=unadj_est,
                       "Standard Error"=unadj_se,
                       "Pvalue"=p_unadj)

p_demo<-ifelse(summary(model_2)$coefficients[,"Pr(>|t|)"]<0.01,"<0.01",round(summary(model_2)$coefficients[,"Pr(>|t|)"],digits=2))
demo_est<-summary(model_2)$coefficients[,"Estimate"]
demo_se<-summary(model_2)$coefficients[,"Std. Error"]
demo_nm<-rownames(summary(model_2)$coefficients)
demo<-data.frame("Variable"=demo_nm,"Estimates"=demo_est,
                 "Standard Error"=demo_se,
                 "Pvalue"=p_demo)
anova(model_2)

p_hth<-ifelse(summary(model_3)$coefficients[,"Pr(>|t|)"]<0.01,"<0.01",round(summary(model_3)$coefficients[,"Pr(>|t|)"],digits=2))
hth_est<-summary(model_3)$coefficients[,"Estimate"]
hth_se<-summary(model_3)$coefficients[,"Std. Error"]
hth_nm<-rownames(summary(model_3)$coefficients)
health<-data.frame("Variable"=hth_nm,"Estimates"=round(hth_est,digits = 2),
                   "Standard Error"=round(hth_se,digits=2),
                   "Pvalue"=p_hth)
anova(model_3)

p_fin<-ifelse(summary(model_final)$coefficients[,"Pr(>|t|)"]<0.01,"<0.01",round(summary(model_final)$coefficients[,"Pr(>|t|)"],digits=2))
fin_est<-summary(model_final)$coefficients[,"Estimate"]
fin_se<-summary(model_final)$coefficients[,"Std. Error"]
fin_nm<-rownames(summary(model_final)$coefficients)
final<-data.frame("Variable"=hth_nm,"Estimates"=hth_est,
                  "Standard Error"=hth_se,
                  "Pvalue"=p_hth)
anova(model_final)
unadjusted
demo
health
final
```
